---
title: "Introduction to Matching Pursuit package"
author: "authors: Artur Gramacki & Jarosław Gramacki"
bibliography: bibliography.bib
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true  
    fig_caption: true
vignette: >
  %\VignetteIndexEntry{Introduction to Matching Pursuit package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>

body {
  background-color: #f7f7f7;
}

.main-container {
  background-color: white;
  padding: 30px;
}

pre {
  background-color: #f4f4f4;
  border: 1px solid #d8d8d8;
  padding: 12px;
  border-radius: 4px;
}

code {
  background-color: #f2f2f2;
  padding: 2px 4px;
  border-radius: 3px;
}

</style>

## R version and vignette generation time
```{r R_version, include = TRUE, echo = FALSE}
out <- version
cat(
  "Generated on: ", 
  format(Sys.time(), "%d-%m-%Y"), 
  "\n", 
  "R version:    ", 
  out$major, ".", 
  out$minor, 
  sep = ""
)
```

## Instalation

```{r instalation, include = TRUE, echo = TRUE, warning = FALSE}
# Install and load required R packages
pkgs <- c(
  "edf", "signal", "RSQLite", "DescTools", "imager", 
  "raster", "graphics", "grDevices", "utils", "knitr", "rmarkdown")

to_install = !pkgs %in% installed.packages()

if(any(to_install)) {
  install.packages(pkgs[to_install], repos = "http://cran.us.r-project.org")
}

library(MatchingPursuit)
```

## Introduction
TODO

@Kunik-2025, @Tropp-2008, @Mallat-1993

```{r Gabor_fun, include = FALSE, echo = FALSE, fig.width = 6, fig.height = 4}
number.of.samples <- 512
sampling.frequency <- 256.0
mean <- 1
phase <- pi
sigma <- 0.5
frequency <- 5.0
normalization = TRUE

out <- Gabor.fun(
  number.of.samples,
  sampling.frequency,
  mean,
  phase,
  sigma,
  frequency,
  normalization
)

plot(out$t, out$gabor, type = "l", xlab = "time [sec.]", ylab = "Gabor fun.", panel.first = grid())
```

## Quick Start {#quick_start}

Generating an example mixing of 7 different non-stationary signals. The resulting signal (shown below in blue) will be used to demonstrate the basic functionality of the package.

```{r 7_non_stationary, include = TRUE, echo = TRUE, warning = FALSE, fig.width = 7, fig.height = 7}
fs <- 1024
T <- 1
t <- seq(0, T - 1 / fs, 1 / fs)
N <- length(t)

# 7 non-stationary signals
x1 <- sin(2 * pi * (10 + 40 * t) * t)                            # linear chirp
x2 <- sin(2 * pi * (20 * t^2) * t)                               # nonlinear chirp
x3 <- (1 + 0.5 * sin(2 * pi * 2 * t))  *  sin(2 * pi * 30 * t)   # AM
x4 <- sin(2 * pi * 50 * t + 5 * sin(2 * pi * 3 * t))             # FM
x5 <- exp(-2 * t)  *  sin(2 * pi * 60 * t)                       # attenuated signal
x6 <- sin(2 * pi * (5 + 20 * sin(2 * pi * t)) * t)               # frequency modulated sine wave
x7 <- t * sin(2 * pi * 40 * t)                                   # increasing amplitude

signal <- data.frame(x = x1 + x2 + x3 + x4 + x5 + x6 + x7)
range <- range(signal)

par(mfrow = c(8, 1), pty = "m")
par(mai = c(0.2, 0.4, 0.1, 0.1))

plot(t, signal$x, type = "l", col = "blue", main = "", xlab = "", ylab = "")
plot(t, x1, type = "l", ylab = "", xlab = "", xaxt = "n", yaxt = "s")
plot(t, x2, type = "l", ylab = "", xlab = "", xaxt = "n", yaxt = "s")
plot(t, x3, type = "l", ylab = "", xlab = "", xaxt = "n", yaxt = "s")
plot(t, x4, type = "l", ylab = "", xlab = "", xaxt = "n", yaxt = "s")
plot(t, x5, type = "l", ylab = "", xlab = "", xaxt = "n", yaxt = "s")
plot(t, x6, type = "l", ylab = "", xlab = "", xaxt = "n", yaxt = "s")
plot(t, x7, type = "l", ylab = "", xlab = "", xaxt = "n", yaxt = "s")
```

**Important note:**
The code to run the `empi.execute()` function has been commented out because executing the `empi.execute()` function downloads an external program from the internet with a very efficient implementation of the Matching Pursuit algorithm (the author calls it *Enhanced Matching Pursuit Implementation*, or *empi* for short, see @Rozanski-2024 for details). The download link and name of the downloaded resource are displayed below.

However, because CRAN's package code verification rules prohibit automatic binary downloads, you do not execute this function here.

When this function is run by the user in interactive mode (the `interactive` function returns `TRUE`), the user is asked if they agree to download the external binary. If they agree, the required files are downloaded and saved in the `r  out <- locate.empi()` `r` tools::file_path_sans_ext(out$fname)` subdirectory of the current directory.

This function generates a SQLite file (`db` extension) that stores all atom parameters. This file was generated by the authors and statically stored in the package. In the next chunk, the `empi2ft()` function uses this file as an input (`db.file` parameter).

```{r locate_empi, include = TRUE, echo = TRUE, warning = FALSE}
locate.empi()
```

```{r empi_execute, include = TRUE, echo = TRUE, warning = FALSE}
# empi.download()

# empi.out <- empi.execute (
#   signal = signal,
#   sampling.rate = 1024,
#   write.to.file = TRUE
# )
```

Generating the final time-frequency map for the selected channel (Fp2_F4). The centers of atoms (in the sense of time and frequency coordinates) are marked with the numbers of subsequent atoms, from the highest to the lowest energy. 

Comparing the two signal twaveforms under the TF map, it can be seen that the original signal and the one after reconstruction differ only minimally.

Below the plot, basic signal parameters are displayed, along with information about the number of atoms into which the input signal was decomposed.

Furthermore, the energy of the input signal and the signal reconstructed from the atoms are also calculated. The obtained result shows that practically all of the energy (and even a little more) of the original signal was "explained" by the generated atoms (100.25%).


```{r sample1, include = TRUE, echo = TRUE, warning = FALSE, fig.width = 7, fig.height = 7}

# Reading a SQLite file where all the generated atom's parameters are stored.
db.file <- system.file("extdata", "sample1.db", package = "MatchingPursuit")

# Create time-frequency map based on MP atoms.
out <- empi2tf(
  db.file = db.file,
  channel = 1,
  mode = "sqrt",
  freq.divide = 4,
  increase.factor= 4,
  display.crosses = FALSE,
  display.atom.numbers = TRUE,
  out.mode = "plot"
)
```

With the `atom.params()` function all the atom parameters can be displayed in a tabular form. Note that the `sample1.db` file stores only one channel. The signal was decomposed into 43 atoms. By analyzing the `position` and `frequency` parameters, you can easily find the so-called blobs corresponding to these atoms in the TF map shown above.

```{r atom_params, include = TRUE, echo = TRUE, warning = FALSE}
file <- system.file("extdata", "sample1.db", package = "MatchingPursuit")
(ap <- atom.params(file))
```

## Electroencephalogram (EEG) analysis

Reading an example EEG signals (EDF file). The signal is 10 seconds long. The sampling rate is 256 Hz. The channels have standard names, which are displayed below. 

```{r eeg_edf_reading, include = TRUE, echo = TRUE, warning = FALSE}
file <- system.file("extdata", "EEG_data_10sec.edf", package = "MatchingPursuit")

# Extract signals from EDF file.
signal <- read.edf.signals(file, resampling = FALSE, from = 0, to = 10)

# Read signal main parameters.
pars <- read.edf.params(file)
sampling.rate <- pars$frequency[1]

# Channel names
pars$channel.name
```

A bipolar montage is created (the classical double banana montage). Each channel compares two adjacent electrodes. In the first step, you define the pairs of electrodes to be connected (`pairs` list). In the second step function `eeg.montage` generates the reqired montage. 

```{r montage, include = TRUE, echo = TRUE, warning = FALSE}
# Pairs of signals for bipolar montage (so called "double banana").
pairs <- list(
  c("Fp2", "F4"),  c("F4",  "C4"),  c("C4",  "P4"),  c("P4",  "O2"),  c("Fp1", "F3"),
  c("F3",  "C3"),  c("C3",  "P3"),  c("P3",  "O1"),  c("Fp2", "F8"),  c("F8",  "T4"),
  c("T4",  "T6"),  c("T6",  "O2"),  c("Fp1", "F7"),  c("F7",  "T3"),  c("T3",  "T5"),
  c("T5",  "O1"),  c("Fz",  "Cz"),  c("Cz",  "Pz"))

# Make the bipolar montage.
bip.montage <- eeg.montage(signal, montage.type = c("bipolar"), bipolar.pairs = pairs)
```

EEG signals are rarely analyzed without first being filtered. Using the `filters.coeff` function, you define filter parameters and then filter the signal. The filter parameters given below are typical values recommended in the literature.

```{r filtering, include = TRUE, echo = TRUE, warning = FALSE}
# Filter parameters that will be used (quite typical in filtering EEG signals).
fc <- filters.coeff(
   fs = sampling.rate,
   notch = c(49, 51),
   lowpass = 40,
   highpass = 1,
)

# Filtering input signals.
bip.montage.filt <- bip.montage 

for (m in 1:ncol(bip.montage)) {
  bip.montage.filt[, m] = signal::filtfilt(fc$bf.notch, bip.montage[, m]);     # 50Hz notch filter
  bip.montage.filt[, m] = signal::filtfilt(fc$bf.low, bip.montage.filt[, m]);  # Low pass IIR Butterworth
  bip.montage.filt[, m] = signal::filtfilt(fc$bf.high, bip.montage.filt[, m]); # High pass IIR Butterwoth
}
```

For the selected channel (after bipolar montage it is `r ch <- 1; colnames(bip.montage.filt)[ch]`), the original and filtered signals are displayed. Clearly, the filtering effect is positive.

```{r no_filtering_and_filtering_plot, include = TRUE, echo = TRUE, warning = FALSE, fig.width = 7, fig.height = 7}

ch <- 1
par(mfrow = c(2, 1), pty = "m")

# Not-filtered signal (raw signal).
plot(
  bip.montage[, ch],
  type = "l",
  panel.first = grid(),
  main = paste(colnames(bip.montage)[ch], " (raw signal)", sep = ""),
  xlab = "sample points",
  ylab = ""
)

# Signal after filtering.
plot(
  bip.montage.filt[, ch],
  type = "l",
  panel.first = grid(),
  main = paste(colnames(bip.montage)[ch], " (filtered signal)", sep = ""),
  xlab = "sample points",
  ylab = ""
)
```

**Important note:** The code below has been commented out. See the explanation given in Section 2. 

```{r eeg_empi_execute, include = TRUE, echo = TRUE, warning = FALSE, fig.width = 7, fig.height = 7}
# empi.download()

# empi.out <- empi.execute (
#   signal = signal,
#   sampling.rate = 1024,
#   write.to.file = TRUE
# )
```

Generating the final time-frequency map for the selected channel (`r ch <- 1; colnames(bip.montage.filt)[ch]`). The centers of atoms (in the sense of time and frequency coordinates) are marked with white crosses. 

Comparing the two signal twaveforms under the TF map, it can be seen that the original signal and the one after reconstruction differ only minimally.

Below the plot, basic signal parameters are displayed, along with information about the number of atoms into which the input signal was decomposed.

Furthermore, the energy of the input signal and the signal reconstructed from the generated atoms are also calculated. The obtained result shows that almost all of the energy of the original signal was “explained” by the generated atoms (97.75%).

```{r eeg_TF, include = TRUE, echo = TRUE, warning = FALSE, fig.width = 7, fig.height = 7}
# Reading a SQLite file where all the generated atom's parameters are stored.
db.file <- system.file("extdata", "EEG_data_10sec_bipolar_filtered.db", package = "MatchingPursuit")

# Generate time-frequency map based on MP atoms.
out <- empi2tf(
  db.file = db.file,
  channel = 1,
  mode = "sqrt",
  freq.divide = 8,
  increase.factor= 4,
  display.crosses = TRUE,
  display.atom.numbers = FALSE,
  out.mode = "plot"
)
```

With the `atom.params()` function all the atom parameters can be displayed in a tabular form. Note that the `EEG_data_10sec_bipolar_filtered.db` file stores 18 channels. To keep the result short, only the first 5 and last 5 atoms are displayed below. The total number of atoms generated is 900 (number of atoms = 50,  number of channels = 18, 50 x 18 = 900).

**Note:** We also note that, in general, the number of atoms generated for each channel need not be the same. However, in our example, we have a special situation where the number of atoms in each channel is exactly the same.

```{r eeg_atom_params, include = TRUE, echo = TRUE, warning = FALSE}
file <- system.file("extdata", "EEG_data_10sec_bipolar_filtered.db", package = "MatchingPursuit")
ap <- atom.params(file)
head(ap, 6)
tail(ap, 6)
```


## Summary
TODO


## Bibliogaphy

