---
title: "Introduction to `MatchingPursuit` package"
author: "authors: Artur Gramacki & Jarosław Gramacki"
bibliography: bibliography.bib
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true  
    fig_caption: true
vignette: >
  %\VignetteIndexEntry{Introduction to `MatchingPursuit` package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>

body {
  background-color: #ffffff;
}

pre {
  border: 1px solid #3b82f6;
  border-radius: 6px;
}

h1, h2, h3, h4 {
  border-bottom: none;
}

</style>

```{r setup, include = TRUE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

## R version and generation time
```{r Setlocale, include = FALSE, echo = FALSE}
Sys.setlocale("LC_ALL", "en_US")
```

```{r R_version, include = TRUE, echo = FALSE}
out <- version
cat(
  "R version:    ", 
  out$major, ".", 
  out$minor, 
  "\n", 
  "Generated on: ", 
  format(Sys.time(), "%d-%B-%Y"), 
  sep = ""
)
```

## Instalation

```{r instalation, include = TRUE, echo = TRUE, warning = FALSE}
# Install and load required R packages
pkgs <- c(
  "edf", "signal", "RSQLite", "DescTools", "imager", "raster", "graphics", "grDevices", 
  "utils", "knitr", "rmarkdown", "latex2exp", "knitr", "rmarkdown", "remotes")

to_install = !pkgs %in% installed.packages()

if(any(to_install)) {
  install.packages(pkgs[to_install], repos = "http://cran.us.r-project.org")
}

library(MatchingPursuit)
```

## The purpose of the package
The presented package allows you to analyse any time-series signals using the Matching Pursuit (MP) algorithm (see @Mallat-1993, @Tropp-2008). Additionally, you can work with EEG electroencephalogram signals by: 

1. Loading files in EDF and EDF+ formats
2. Performing 3 typical EEG montages
3. Pre-filtering the signals


## Quick start 

Let's generate an example mixing of seven different non-stationary signals. The resulting signal (highlighted in blue below) will be used to demonstrate the basic functionality of the package.

**STEP 1:**

```{r 7_non_stationary, include = TRUE, echo = FALSE, warning = FALSE, fig.width = 7, fig.height = 7}
fs <- 1024
T <- 1
t <- seq(0, T - 1 / fs, 1 / fs)
N <- length(t)

# 7 non-stationary signals
x1 <- sin(2 * pi * (10 + 40 * t) * t)                            # linear chirp
x2 <- sin(2 * pi * (20 * t^2) * t)                               # nonlinear chirp
x3 <- (1 + 0.5 * sin(2 * pi * 2 * t))  *  sin(2 * pi * 30 * t)   # AM
x4 <- sin(2 * pi * 50 * t + 5 * sin(2 * pi * 3 * t))             # FM
x5 <- exp(-2 * t)  *  sin(2 * pi * 60 * t)                       # decreasing amplitude
x6 <- sin(2 * pi * (5 + 20 * sin(2 * pi * t)) * t)               # frequency modulated sine wave
x7 <- t * sin(2 * pi * 40 * t)                                   # increasing amplitude

signal <- data.frame(x = x1 + x2 + x3 + x4 + x5 + x6 + x7)
range <- range(signal)

par(mfrow = c(8, 1), pty = "m")
par(mai = c(0.2, 0.4, 0.2, 0.1))

plot(t, signal$x, type = "l", col = "blue", main = "", xlab = "", ylab = "")
plot(t, x1, type = "l", ylab = "", xlab = "", xaxt = "n", yaxt = "s", main = "linear chirp")
plot(t, x2, type = "l", ylab = "", xlab = "", xaxt = "n", yaxt = "s", main = "nonlinear chirp")
plot(t, x3, type = "l", ylab = "", xlab = "", xaxt = "n", yaxt = "s", main = "AM")
plot(t, x4, type = "l", ylab = "", xlab = "", xaxt = "n", yaxt = "s", main = "FM")
plot(t, x5, type = "l", ylab = "", xlab = "", xaxt = "n", yaxt = "s", main = "decreasing amplitude")
plot(t, x6, type = "l", ylab = "", xlab = "", xaxt = "n", yaxt = "s", main = "frequency modulated sine wave")
plot(t, x7, type = "l", ylab = "", xlab = "", xaxt = "n", yaxt = "s", main = "increasing amplitude")
```

Data must my stored in data frame: rows = samples for all channels, columns = channels. Our first demo dataset consists of only one channel (one column).

```{r signal, include = TRUE, echo = TRUE, warning = FALSE}
file <- system.file("extdata", "sample1.csv", package = "MatchingPursuit")
signal <- read.csv(file, header = TRUE)

class(signal)

head(signal)
```

**STEP 2:**

The input data (`signal`) is passed as an input parameter to the `empi.execute()` function which generates a file in `SQLite` format (`empi.db`) that stores all atom parameters. This `.db` file is stored in a temporary directory created by `tempfile()` function.

The function first downloads (if it hasn't been downloaded before) and then runs an external program from the internet. This program is a very fast and efficient implementation of the Matching Pursuit algorithm. The author calls it **Enhanced Matching Pursuit Implementation**, or **empi** for short, see @Rozanski-2024 for details. The download link and name of the downloaded resource are displayed below (if running in Linux and macOS systems, appropriate binary versions are downloaded).

```{r locate_empi, include = TRUE, echo = TRUE, warning = FALSE}
locate.empi()
```

<div id="important_note"></div>
**Important note:**
The code in the below chunk has been commented out because CRAN's code verification rules prevent automatic binary downloads. Therefore, this function should not be run here. However, when this function is run by the user in interactive mode (if the `interactive()` function returns `TRUE`), the user is asked if they agree to download the external binary. If they agree, the required files are downloaded and saved in the current directory.


Due to CRAN limitations (discussed above), this file was generated by the authors and stored in the package (`sample1.db`) . In the next chunk, the `empi2ft()` function uses this file as input (the `db.file` parameter).

Notice the `empi.options` parameter in the `empi.execute()` function. You can specify the value `NULL` there, and *empi* program will run with the default values  set in the function (`"-o local --gabor -i 50"`). It is also worth noting that it has a wide range of configuration options. Details can be found in the `README.md` (in the directory where the *empi* program was saved) file. In our example, parameters were entered that tell the program to find 25 atoms.

```{r empi_execute, include = TRUE, echo = TRUE, warning = FALSE}
# empi.download()

# empi.out <- empi.execute (
#   signal = signal,
#   sampling.rate = 1024,
#   empi.options = "-o local --gabor -i 25",
#   write.to.file = TRUE
# )
```

**STEP 3:**

Generating the final time-frequency map (T-F) for the selected channel. The centres of atoms (in the sense of time and frequency coordinates) are marked with the numbers of subsequent atoms, sorted from the highest to the lowest energy. 

Comparing the two signal waveforms under the T-F map, it can be seen that the original signal and the one after reconstruction differ only minimally.

```{r sample1, include = TRUE, echo = TRUE, warning = FALSE, fig.width = 7, fig.height = 7}
# Reading a SQLite file where all the generated atom's parameters are stored.
db.file <- system.file("extdata", "sample1.db", package = "MatchingPursuit")

# Create time-frequency map based on MP atoms.
out <- empi2tf(
  db.file = db.file,
  channel = 1,
  mode = "sqrt",
  freq.divide = 4,
  increase.factor= 4,
  display.crosses = FALSE,
  display.atom.numbers = TRUE,
  out.mode = "plot"
)
```

Below the plot, basic signal parameters are displayed, along with information about the number of atoms into which the input signal was decomposed.

Furthermore, the energy of the input signal and the signal reconstructed from the atoms are also calculated. The obtained result shows that more of the energy of the original signal was "explained" by the generated atoms (94.05%). Of course, the number of generated atoms can be increased, then the energy of the reconstructed signal will most likely be closer to 100%.

**STEP 4 (not obligatory):**

With the `atom.params()` function all the atom parameters can be displayed in a tabular form. Note that the `sample1.db` file stores only one channel. The signal was decomposed into 43 atoms. By analysing the `position` and `frequency` parameters, you can easily find the so-called blobs corresponding to these atoms in the T-F map shown above.

```{r atom_params, include = TRUE, echo = TRUE, warning = FALSE}
file <- system.file("extdata", "sample1.db", package = "MatchingPursuit")
(ap <- atom.params(file))
```

**STEP 5 (not obligatory):**

The `sig2bin()` function can be used to write input data to a binary file. This function is is used internally in the `empi.execute()` function. The binary data are floating-point values in the byte order of the current machine (no byte-order conversion is performed). For multichannel signals, first come the samples for all channels at `t = 0`, then for all channels at `t = Δt`, and so forth. In other words, the signal should be written in column-major order (rows = channels, columns = samples).

**Note:** The user does not work directly with `.bin` files. Binary files are used only in the `empi.execute()` function. The external **empi** program executed inside this function requires binary data as input. Moreover, the ability to convert text files to binary form may be useful if someone wants to work with **empi** independently of the R environment.

```{r sig_2_bin, include = TRUE, echo = TRUE, warning = FALSE}
signal.bin <- sig2bin(data = signal, write.to.file = FALSE)
class(signal.bin)

# First 4 elements of the signal.
head(signal, 4)

# The same first 4 elements of the signal in binary (float, 4 bytes).
head(signal.bin, 16)
```

**STEP 6 (not obligatory):**

Data from the generated `SQLite` file can be easily read using the `read.empi.db.file()` function. Below, data from a sample `.db` file is read. Original and the reconstructed signals are then extracted and displayed. The two signals are virtually indistinguishable from each other. This means that atoms are very good at reconstructing the analysed signal.

```{r read_empi_db_file, include = TRUE, echo = TRUE, warning = FALSE}
file <- system.file("extdata", "sample1.db", package = "MatchingPursuit")
out <- read.empi.db.file(file)
```

```{r display_original_and_reconstrction, include = TRUE, echo = FALSE, warning = FALSE, fig.width = 7, fig.height = 4}
n.channnels <- ncol(out$original.signal)
original.signal <- out$original.signal
reconstruction <- out$reconstruction
t <- out$t
f <- out$f
len <- length(original.signal[, 1])
lab <- seq(t[1], t[len] + 1 / f, length.out = 11)

par(mfrow = c(2, 1), pty = "m")
par(mai = c(0.8, 0.5, 0.5, 0.5))

plot(
  original.signal[,1], type = "l", col = "blue",
  main = "Original signal",
  xaxt = "n", ylab = "", xlab = "time [sec]"
)
axis(side = 1, las = 1, cex.axis = 0.9, at = seq(0, len, length.out = 11), labels = lab)

plot(
  reconstruction[,1], type = "l", col = "blue",
  main = "Rconstructed signal",
  xaxt = "n", ylab = "", xlab = "time [sec]"
)
axis(side = 1, las = 1, cex.axis = 0.9, at = seq(0, len, length.out = 11), labels = lab)
```

## Electroencephalogram (EEG) analysis

In this section, we demonstrate how to analyze electroencephalography (EEG) signals using the Matching Pursuit algorithm. The package includes a dedicated function for reading files saved in EDF and EDF+ formats (European Data Format). You can also perform three different types of montages as well as filtering the signals.

**STEP 1:**

Reading an example EEG signals (EDF file). The signal is 10 seconds long and consists of 20 channels.e The sampling rate is 256 Hz. The channels have standard names, which are displayed below. 

```{r eeg_edf_reading, include = TRUE, echo = TRUE, warning = FALSE}
file <- system.file("extdata", "EEG.edf", package = "MatchingPursuit")

# Extract signals from EDF file.
signal <- read.edf.signals(file, resampling = FALSE, from = 0, to = 10)

# Read signal main parameters.
pars <- read.edf.params(file)
sampling.rate <- pars$frequency[1]

# Read signal parameters and display them in a tabular form.
read.edf.params(file)
```

**STEP 2 (sometimes required):**

Sometimes it is necessary to change the sampling frequency (increase -- *upsampling* or decrease -- *downsampling*). The `read.edf.signals()` function implements this functionality. In the example below the original sampling frequency is reduced from 256 Hz to 64 Hz.

```{r eeg_edf_reading_resampling, include = TRUE, echo = TRUE, warning = FALSE}
file <- system.file("extdata", "EEG.edf", package = "MatchingPursuit")

# Original signal
signal <- read.edf.signals(file, resampling = FALSE, from = 0, to = 10)

# Resampled signal
signal.resampled <- read.edf.signals(file, resampling = TRUE, f.new = 64,  from = 0, to = 10)
```

```{r sampling, include = TRUE, echo = FALSE, warning = FALSE, fig.width = 7, fig.height = 4}
par(mfrow = c(2, 1), pty = "m")
par(mai = c(0.8, 0.5, 0.5, 0.5))

# Not-filtered signal (raw signal).
plot(
  signal[, 1],
  type = "l",
  panel.first = grid(),
  main = "Original signal (256 Hz)",
  xlab = "sample points",
  ylab = "",
  col = "blue"
)

# Signal after filtering.
plot(
  signal.resampled[, 1],
  type = "l",
  panel.first = grid(),
  main = "Signal after downsampling (64 Hz)",
  xlab = "sample points",
  ylab = "",
  col = "blue"
)
```


**STEP 3 (usually required):**

A bipolar montage is created (the classical *double banana* montage). Each channel compares two adjacent electrodes. In the first step, you define the pairs of electrodes to be connected (`pairs` list). In the second step function `eeg.montage()` generates the required montage. 

```{r montage, include = TRUE, echo = TRUE, warning = FALSE}
# Pairs of signals for bipolar montage (so called "double banana").
pairs <- list(
  c("Fp2", "F4"), c("F4",  "C4"), c("C4",  "P4"), c("P4",  "O2"), c("Fp1", "F3"), c("F3",  "C3"),  
  c("C3",  "P3"), c("P3",  "O1"), c("Fp2", "F8"), c("F8",  "T4"), c("T4",  "T6"), c("T6",  "O2"),
  c("Fp1", "F7"), c("F7",  "T3"), c("T3",  "T5"), c("T5",  "O1"), c("Fz",  "Cz"), c("Cz",  "Pz")
)

# Make the bipolar montage.
bip.montage <- eeg.montage(signal, montage.type = c("bipolar"), bipolar.pairs = pairs)

# Original signal.
head(signal[, 1:7])

# Signal after banana montage.
head(bip.montage[, 1:7])

```

**STEP 4 (usually required):**

EEG signals are rarely analysed without first being filtered. Using the `filters.coeff()` function, you define filter parameters and then filter the signal. The filter parameters given below have typical values recommended in the literature.

```{r filtering, include = TRUE, echo = TRUE, warning = FALSE}
# Filter parameters that will be used (quite typical in filtering EEG signals).
fc <- filters.coeff(
   fs = sampling.rate,
   notch = c(49, 51),
   lowpass = 40,
   highpass = 1,
)

# Filtering input signals.
bip.montage.filt <- bip.montage 

for (m in 1:ncol(bip.montage)) {
  bip.montage.filt[, m] = signal::filtfilt(fc$notch, bip.montage[, m])     # 50Hz notch filter
  bip.montage.filt[, m] = signal::filtfilt(fc$low, bip.montage.filt[, m])  # Low pass IIR Butterworth
  bip.montage.filt[, m] = signal::filtfilt(fc$high, bip.montage.filt[, m]) # High pass IIR Butterwoth
}
```

For the selected channel (after bipolar montage it is named `r ch <- 1; colnames(bip.montage.filt)[ch]`), the original and filtered signals are displayed. Clearly, the filtering effect is positive.

```{r no_filtering_and_filtering_plot, include = TRUE, echo = FALSE, warning = FALSE, fig.width = 7, fig.height = 4}
ch <- 1
par(mfrow = c(2, 1), pty = "m")
par(mai = c(0.8, 0.5, 0.5, 0.5))

# Not-filtered signal (raw signal).
plot(
  bip.montage[, ch],
  type = "l",
  panel.first = grid(),
  main = paste(colnames(bip.montage)[ch], " (raw signal)", sep = ""),
  xlab = "sample points",
  ylab = "",
  col = "blue"
)

# Signal after filtering.
plot(
  bip.montage.filt[, ch],
  type = "l",
  panel.first = grid(),
  main = paste(colnames(bip.montage)[ch], " (filtered signal)", sep = ""),
  xlab = "sample points",
  ylab = "",
  col = "blue"
)
```

**STEP 5:**

**Important note:** The code below has been commented out. See the explanation given [here](#important_note).

```{r eeg_empi_execute, include = TRUE, echo = TRUE, warning = FALSE, fig.width = 7, fig.height = 7}
# empi.download()

# The empi.options parameter is NULL, so the empi program is 
# run with the parameters "-o local --gabor -i 50"
# empi.out <- empi.execute (
#   signal = signal,
#   sampling.rate = 1024,
#   empi.options = NULL,
#   write.to.file = TRUE
# )
```

Generating the final time-frequency map for the selected channel (`r ch <- 1; colnames(bip.montage.filt)[ch]`). The canters of atoms (in the sense of time and frequency coordinates) are marked with white crosses. 

Comparing the two signal twaveforms under the T-F map, it can be seen that the original signal and the one after reconstruction differ only minimally.

```{r eeg_TF, include = TRUE, echo = TRUE, warning = FALSE, fig.width = 7, fig.height = 7}
# Reading a SQLite file where all the generated atom's parameters are stored.
db.file <- system.file("extdata", "EEG_bipolar_filtered.db", package = "MatchingPursuit")

# Generate time-frequency map based on MP atoms.
out <- empi2tf(
  db.file = db.file,
  channel = 1,
  mode = "sqrt",
  freq.divide = 6,
  increase.factor= 4,
  display.crosses = TRUE,
  display.atom.numbers = FALSE,
  out.mode = "plot"
)
```

Below the plot, basic signal parameters are displayed, along with information about the number of atoms into which the input signal was decomposed.

Furthermore, the energy of the input signal and the signal reconstructed from the generated atoms are also calculated. The obtained result shows that almost all of the energy of the original signal was “explained” by the generated atoms (97.75%).

**STEP 6 (not obligatory):**

With the `atom.params()` function all the atom parameters can be displayed in a tabular form. Note that the `EEG_bipolar_filtered.db` file stores 18 channels. To keep the result short, only the first 5 and last 5 atoms are displayed below. The total number of atoms generated is 900 (number of atoms = 50,  number of channels = 18, 50 x 18 = 900).

**Important:** It should be clearly emphasized that in general, the number of atoms generated for each channel need not be the same. However, in our example, we have a special situation where the number of atoms in each channel is exactly the same.

```{r eeg_atom_params, include = TRUE, echo = TRUE, warning = FALSE}
file <- system.file("extdata", "EEG_bipolar_filtered.db", package = "MatchingPursuit")
ap <- atom.params(file)
head(ap, 6)
tail(ap, 6)
```

## The Matching Pursuit algorithm in a nutshell
The Matching Pursuit algorithm is well-known and described in the literature Its purpose is to approximate the analyzed signal using so-called atoms. (the text below is adopted from @Kunik-2025).

Given a signal $f \in \mathbb{R}^n$, and a (possibly overcomplete) large redundant dictionary $D =\{g_{\gamma}\}_{\gamma \in \Gamma}$ of normalized atoms $\|g_{\gamma}\|=1$ MP finds a sparse signal representation

$$
f \approx \sum_{n = 0}^{N} a_n g_{\gamma_n}
\tag{1}
$$
where $a_n \in \mathbb{R}$ are coefficients, $g_{\gamma_n} \in D$ are atoms selected from the dictionary and $N$ is the desired number of iterations (or stopping threshold). In most practical cases $ N \ll size(D)$. Also, $g_{\gamma}$ is the dictionary atom indexed by $\gamma$ and $\Gamma$ is the set of all indices in the directory.

In the ideal situation linear expansion (1) should contain all the atoms $g_{\gamma_n}$ that represent relevant structures of the signal $f$. In real signals, such an ideal situation is rarely possible and a kind of approximation is needed. This task can be performed in a very elegant way using the MP algorithm. This algorithm was first proposed by @Mallat-1993 in the context of signal processing. 

Each atom $g_{\gamma}$ is typically a time-frequency shifted and scaled version of a prototype function, such as the Gabor function (often called a Gaussian-windowed sinusoid). The dictionary is constructed to span a wide range of time and frequency characteristics. Real-valued Gabor function has the following form

$$
g_{\gamma}(t) = 
K(\gamma) e^{- \pi \left( \frac{t-\mu}{\sigma} \right) ^2}
\cos(\omega (t - \mu) + \phi)
$$

where $\gamma = (\mu, \omega, \sigma, \phi)$ constitute a four-dimmensional space and $K(\gamma)$ is such that $||g_{\gamma}|| = 1$. It is easy to see that Gabor functions are constructed by multiplying Gaussian envelopes with cosine oscillations of different frequencies $\omega$ and phases offset $\phi$. By multiplying these two functions, we can obtain a wide variety of shapes depending on their parameters. A few examples of Gabor function are presented in figure below (in blue). The sinusoidal plane wave (in gray) is modulated by a Gaussian envelope (in red).

```{r Gabor_fun, include = TRUE, echo = FALSE, fig.width = 7, fig.height = 5, fig.align = 'center'}
library(latex2exp)
N <- 512
fs <- 256
# normalization = T --> signal's norm = 1

par(mfrow = c(2,2), pty="m")
par(mai=c(0.4, 0.4, 0.3, 0.2))

n <- 4

sigmas <- c(0.5, 0.2, 0.8, 0.5)
frequencies <- c(14, 8, 4, 1)
phases <- c(0, 1, 1.5, -2)
means = c(0.5, 0.8, 1, 1.5)

s <- rep(0, N)

for (i in c(1:n)) {
  sigma <- sigmas[i]
  frequency <- frequencies[i]
  phase <- phases[i]
  mean <- means[i]
  main <- TeX(paste(
    "$\\mu=$", means[i], ", ", 
    "$\\sigma=$", sigmas[i], ", ",
    "$\\f=$", frequencies[i], ", ",
    "$\\phi=$", phases[i], 
    sep = ""
  )
  )
  
  gb <- Gabor.fun(N, fs, mean, phase, sigma, frequency, normalization = F)

  plot(
    gb$t, 
    gb$gauss, type="l", ylim = c(-1, 1), col = "red", 
    xlab= "", ylab= "",
    xaxt = "t", yaxt = "t", bty = "or",
    cex.axis = 1, lwd = 2,
    main = main)
  lines(gb$t, gb$cosinus, type = "l", col = "grey")
  lines(gb$t, gb$gabor, col="blue", lwd = 2)
  
  s <- s + gb$gabor
}
```

In the MP algorithm, the decomposition process is iterative. At each step, the algorithm selects an atom $g_{\gamma_n}$ from the dictionary $D$ that best matches the current residual signal $R$.
Formally, starting with the signal $f$ at iteration $n=0$ the initial residual and initial function approximation are

$$
R^0 = f \\
f^0 = 0
$$

For each iteration $n = \{0,1,\ldots, N\}$ we find $g_{\gamma_n} \in D$ such that the following inner product $\langle \cdot, \cdot \rangle$ is maximized

$$
g_{\gamma_n} = \arg \max_{\gamma \in \Gamma} | \langle R^{n}, g_{\gamma} \rangle |
$$

The coefficients $a_n$ in (1) are 

$$
a_n = \langle R^{n}, g_{\gamma_n} \rangle 
$$

and updated function approximation is defined as

$$
f^{n+1} = f^{n} + a_n g_{\gamma_n}
$$
Similarly, updated residual is defined as

$$
R^{n+1} = R^{n} - a_n g_{\gamma_n}
$$
After $N$ iterations, the signal $f$ is approximated as


$$
f \approx \sum_{n = 0}^{N} \langle R^n,  g_{\gamma_n} \rangle g_{\gamma_n} = \sum_{n = 0}^{N} a_n g_{\gamma_n}
$$

or equivalently

$$
f = \sum_{n = 0}^{N} \langle R^n   g_{\gamma_n} \rangle g_{\gamma_n} + R^{N+1}
$$

The procedure stops when $\|R^{n+1}\|$ is below a threshold or after a fixed number of iterations.

It should be mentioned here that finding an optimal approximation (1) is an NP-hard problem. A suboptimal solution can be found by means of an iterative procedure, such as the MP algorithm. Also, a key property of MP is energy conservation. The total energy of the signal is preserved in the MP decomposition


$$
||f||^2 = \sum_{n = 0}^{N} |\langle R^n,  g_{\gamma_n} \rangle |^2 + ||R^{N+1}||^2
$$

## Bibliogaphy

